---
title: "README"
output: html_document
---

Aim of this folder is to explore the Hierarchial Risk Partiy approach to portfolio optimization.

Paper: https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2708678

R Code walk through: http://gallery.rcpp.org/articles/hierarchical-risk-parity/


```{r}
# Ensure the following packages are installed ====================
install.packages("tidyverse")
install.packages("PerformanceAnalytics")
install.packages("devtools")
library(devtools)
install_github("Nicktz/Dplyr2Xts")
install.packages("Rcpp")
install.packages("RcppArmadillo")

```


```{r}

library(tidyverse)
library(PerformanceAnalytics)
library(Dplyr2Xts)
library(Rcpp)
library(RcppArmadillo)
Root <- getwd()
# Load Cpp files:====================
l <- list.files(paste0(Root, "/Code"), full.names = TRUE )
sapply(FUN = Rcpp::sourceCpp, X = l)

```



# Step 1: Load Toy returns data

```{r}
data("edhec")
datwide <- edhec %>% Xts2Dplyr()

CorMat <- 
  cor(datwide[-1], use = "pairwise.complete.obs") 
CovMat <- 
  cov(datwide[-1], use = "pairwise.complete.obs") 


```


Now see this set of slides explaining the process more easily: 
https://poseidon01.ssrn.com/delivery.php?ID=712085084074106099027084030066071002122081004009095091026103081078100115069009007026018029054101050127023023094092101125003071058082046034028110102099101019098010039021095117067125006091127091121115107029112090072118067092005077096024110110024106115&EXT=pdf


# Step 2: Rowwise and Elementwise distance matrix calculation:


```{r}

distanceElementWise <- distanceMatrix_elementwise(CorMat)

distanceRowwise <- distanceMatrix_rowwise(CorMat)

```



# Step 3: Cluster Generation

```{r}

Cl <- clusterMatrix(CorMat)

```



# Step 4: Quasi-Diagonalization


```{r}

CLIndex <- clusterIndex(clusterMatrix = Cl, MAT_COV = CovMat)

QuasiDiagMatrix <- quasiDiag(MAT_COV = CovMat, assetIndex = CLIndex)

```


# Step 5: Weighting Generation


```{r}


WeightVector <- weightAllocation(quasiDiagMatrix = QuasiDiagMatrix, assetIndex = CLIndex)

colnames(WeightVector) <- TickerstoUse

WeightVector %>% tbl_df() %>% gather(Ticker, Weight) %>% arrange(desc(Weight))

```

### Note: This Weightvector is as of yet unconstrained...

TODO:

Build a more flexible package in R to do HRP optimization.



